---
title: "Quantified Self Movement Prediction"
author: "Bill Holst"
date: "April 12, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

This project examines a dataset from a group that studies human motion through measurments from accelerometers in fitness devices. 

The goal of this project is to predict the manner in which particpants did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.
```{r cars}
summary(cars)
```

## Posing the Question

uses impute...
https://github.com/jasoniem9246/Prediction-Assignment-Writeup/blob/master/Prediction%20Assignment%20Writeup.Rmd

good writeup - https://rpubs.com/Shikin2015/Prediction_Assignment


different models - shows out of model error

https://rpubs.com/snowdj/pml
You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


testing = adData[-trainIndex,]

adData = data.frame(diagnosis,predictors)
trainIndex = createDataPartition(diagnosis, p = 0.50)
training = adData[trainIndex,]
testing = adData[-trainIndex,]

adData = data.frame(predictors)
trainIndex = createDataPartition(diagnosis,p=0.5,list=FALSE)
training = adData[trainIndex,]
testing = adData[-trainIndex,]

trainingIL <- training[, c(ILpredictor, "diagnosis")]
testingIL <- testing[, c(ILpredictor, "diagnosis")]
ModelAll <- train(diagnosis ~ ., data = trainingIL, method = "glm")
confusionMatrix(testingIL$diagnosis, predict(ModelAll, testingIL))

preProc <- preProcess(training[, ILpredictor], method = "pca", thresh = .8)
trainPC <- predict(preProc, training[, ILpredictor])
ModelPCA <- train(trainingIL$diagnosis ~ ., method = "glm", data = trainPC)
testPC <- predict(preProc, testing[, ILpredictor])
confusionMatrix(testingIL$diagnosis, predict(ModelPCA, testPC))

## Data Analysis

This section describes the dataset and explains how the data was obtained, how predictors are established, how parameters are picked, and algorithms evaluated.

### Getting the Data

First we need to download and read the data.

URLtrain = 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
training = read.csv (URLtrain, na.strings=c("NA","#DIV/0!",""))



urltest = 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'
testing = read.csv (urltest,na.strings=c("NA","#DIV/0!",""))

A quick look at the summary shows that many of the variables are sparsely populated, with 19k NA values. We eliminate these.

tnum = NULL

for (i in 1:length(training)) {
  if (sum(is.na(training[,i]))<=19000) {
    tnum = cbind(tnum,i) 
  }
}
# remove the na columns
newtraining = training[,c(tnum)]

We can also remove the near zero variance variables.

nz = nearZeroVar(newtraining,saveMetrics = TRUE)


### Predictors, Parameters, and Algorithms

Because the intended 

## Prediction

## Summary

## References

Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6. 

Read more: http://groupware.les.inf.puc-rio.br/har#sbia_paper_section#ixzz46fnQWmUP
